When does vapor pressure deficit drive or reduce evapotranspiration?

### 2019-02-27 doc/paper/abstract.tex
about to start responding to pierre's comments

The ET response varies due to: 1) climate, with tropical and
temperate climates more likely to exhibit a positive ET response than
boreal and arctic climates; 2) photosynthesis strategy, with C3 plants
more likely to exhibit a positive ET response than C4 plants; and 3)
due to plant type, with crops more likely to exhibit a positive ET
response, and shrubs and gymniosperm trees more likely to exhibit a
negative ET response. These results, derived from previous literature
connecting plant parameters to plant and climate characteristics,
highlight the utility of our simplified framework for understanding
complex land atmosphere systems in terms of idealized scenarios.

>>>>

The ET response varies due to: 1) climate, with tropical and temperate
climates more likely to exhibit a positive ET response to increasing
VPD than boreal and arctic climates; 2) photosynthesis strategy, with
C3 plants more likely to exhibit a positive ET response than C4
plants; and 3) due to plant type, with crops more likely to exhibit a
positive ET response, and shrubs and gymniosperm trees more likely to
exhibit a negative ET response. These results, derived from previous
literature connecting plant parameters to plant and climate
characteristics, highlight the utility of our simplified framework for
understanding complex land atmosphere systems in terms of idealized
scenarios.

### 2019-02-27 doc/paper/body.tex
about to start responding to pierre's comments

The Medlyn model for stomatal conductance has been shown to behave
very well across PFTs \citep[][]{Lin_2015}, and has been successfully
adopted to ecosystem scale analysis in \citet{Medlyn_2017}. In units
of m s$^{-1}$, the ecosystem scale stomatal conductance is:

@@

  where GPP is the ecosystem scale gross primary production, and $g_1$
  is an ecosystem scale analogue to $g_{1-leaf}$. We solve for $g_1$
  following \citet{Medlyn_2017} (Eq. (5)), and take the median $g_1$
  value to be representative of each PFT (Table \ref{pft}) instead of
  the mean to avoid extra weighting of rare outliers.

While Eq. (3) can be used in PM (Eq. (1)), it will make
analytical work with the function intractable because $GPP$ is
functionally related to ET itself. Additionally, a perturbation to VPD
should induce a physiological plant response that will alter GPP and
cause an indirect change in stomatal conductance, in addition to the
direct effect of VPD in Eq. (\ref{medlyn}). Therefore, in order to
derive the response of ET to VPD, we must account for the functional
relationship between GPP, ET, and VPD, and its effect on stomatal
conductance. We can use aforementioned semi-empirical results of
\citet{Zhou_2015} as a tool to approach this
problem. \citet{Zhou_2015}, showed that  underlying Water Use
Efficiency (uWUE):

@@

is relatively constant across time and moisture conditions within a
plant functional type, and correctly captures a constant relationship
between GPP, ET and VPD over a diurnal cycle \citep{Zhou_2014}. The
theoretical derivation of the square root VPD dependence in $uWUE$
leverages the same assumptions used in \cite{MEDLYN_2011} to derive the
square-root VPD dependence of the stomatal conductance model (Eq.
(\ref{medlyn})).  We can use uWUE to remove the $GPP$ dependence from
$g_s$ in a way that makes PM analytically tractable:

@@

With Equation \ref{eq} we can take the partial derivative with respect
to VPD:

@@

  This derivation relied either implicitly or explicitly on several
  assumptions. First, we assume that VPD at the leaf surface is the
  same as VPD at measurement height; physically this implies that
  leaves are perfectly coupled to the atmosphere. In reality, for some
  conditions and plant types the leaves can become decoupled from the
  boundary layer \citep{De_2017, Medlyn_2017}. Therefor, our
  derivation will be most applicable in times like the growing season,
  when relatively high insolation induces instability and convective
  boundary layers, and we would expect the surface to be generally
  well coupled. An additional assumption in the formulation of $uWUE$
  \citep{Zhou_2014, Zhou_2015} and \citet{Medlyn_2017}'s stomatal
  conductance model is that direct soil evaporation (E) contributions
  to ET remain small relative to transpiration (T). This should be
  more true during the growing season. The ratio of E to T may
  increase immediately after rainfall events due to high soil moisture
  and ponding, but VPD is generally low anyways during these
  times. However, some plant types allow for systematically larger
  contributions of E in ET, particularly those with sparse canopies
  and smaller relative amounts of transpiration. We therefore might
  expect that the theory will be most applicable to forest PFTs, which
  will be most strongly coupled to the boundary layer due to larger
  surface roughness, and will also generally have the highest ratios
  of transpiration to evaporation. Finally, we assume that $g_1$ and
  $uWUE$ are constant with respect to the conceptual VPD
  perturbation. Both quantities have been shown to be relatively
  constant with respect to changes in
  % TODO: make this a general discussion on partial derivatives and
  % what things are constant/how to interpret?
  VPD \cite{Franks_2017, Zhou_2014}. These parameters will however
  vary with plant species and characteristics \citep[e.g. wood density,
  ]{Lin_2015}, as well as environmental conditions including soil
  water content and temperature.
  % TODO: write exact parameter ranges we use, and introduce
  % scaling/sign term here?
  % \item

  To account for both this variability in plant parameters and the
  environment, we will systematically analyze how the ET response to
  VPD (Equation \ref{d_et}) varies. Equation \ref{d_et} includes a
  ``sign'' term that determines the sign of the response in addition
  to magnitude:

@@

and a ``scaling'' multiplying the ``sign'' term:

@@

In the ``sign term'' most of quantities are relatively constant,
except for VPD and the plant parameters g$_1$ and uWUE. In the scaling
term, most of the terms are relatively constant with the exception of
aerodynamic conductance ($g_a$) and temperature (especially its effect
through $\Delta$). To determine the range of probable ET responses to
VPD we will systematically vary this parameters according to Table
\ref{param_varying}, while all other parameters are held fixed (Table
\ref{param_fixed}). Physical variables ($g_a$, $T$) and plant
physiological parameters (g$_1$, uWUE) are varied according to
literature-based expectations for a range of growing season conditions
and plant types \citep{Zhou_2015, Medlyn_2017}. Using this previous
literature we can connect the effect of varying plant parameters to
specific plant types and characteristics.

@@

\begin{table}
  \caption{Variable quantities in the ET response to VPD. Each value
    is varied to determine the effect of a range of expected plant and
    environmental conditions on ET response to VPD. A citation is
    provided in cases where the quantities are directly derived from
    previous literature. Conceptual percentiles
    for interpreting each quantity are noted.}
  \label{param_varying}
  \centering
  \begin{tabular}{l c c c c c}
    \hline
    Symbol & units [units in citation] & min (~15th percentile) & med (~50th percentile) & max
                                                                                            (~15th
                                                                                            percentile)
    & citation  \\
    \hline
    \input{param_varying.tex}
    \hline
  \end{tabular}
\end{table}

\begin{table}
  \caption{Quantities that are fixed in the ET response to VPD
    (relative to those in Table \ref{param_varying}).}
  \label{param_varying}
  \centering
  \begin{tabular}{l c c}
    \hline
    Symbol & units & value \\
    \hline
    \input{param_fixed.tex}
    \hline
  \end{tabular}
\end{table}

@@

\begin{figure}
  \centering \includegraphics{./fully_idealized.pdf}
  \caption{The functional form of the $\frac{\partial \; ET}{\partial
      \; VPD}$ for minimum, median and maximum values of both the sign
    term and the scaling term.}
  \label{full}
\end{figure}

Examining the sign term independent of the scaling term indicates the
dependence of water conservative versus water intensive response on
the main plant parameters, g$_1$ and uWUE (Figure \ref{sign}). Higher
g$_1$ and uWUE shift the curve towards increasing ET responses with
VPD (water intensive), and with increasing g$_1$ also leading to a
higher non-linear VPD dependence of the response.

\begin{figure}
  \centering \includegraphics{./fully_idealized_sign.pdf}
  \caption{The functional form of the $\frac{\partial \; ET}{\partial
      \; VPD}$ evaluated at the median value of the scaling term, for
    varying values of g$_1$ and uWUE as given in Table \ref{param_varying}.}
  \label{sign}
\end{figure}

Examining the scaling term independent of the sign term shows how
aerodynamic conductance (communication between the atmosphere and the
surface) and temperature (through $\Delta$ controls the efficiency of
energy conversion to latent heat), amplify or suppress the plant
physiological response represented in the sign term. Both lower
temperatures and higher aerodynamic conductance lead to amplified ET
response to VPD, with the variability of aerodynamic conductance
resulting in a slightly higher variability of ET response relative to
temperature, which is still an important factor.

@@

Interpreting both the derivation and results of this manuscript hinges
on an understanding and appreciation for the usefulness of partial
derivatives for understanding the behavior of complex systems. Our
results are a conceptual tool for answering the question, ``What is
the ET response to VPD with all other quantities held fixed?'' This is
useful, and we would argue is critical, for understanding complex
responses when many quantities are varying simultaneously. Through
partial derivatives, we can approach answering questions like ``What
is the relative contribution of VPD in some observed response given
that X Y and Z were also varying?'' In this example, X Y and Z could
be, for example, radiation, soil moisture, and temperature. Equation
\ref{d_et} explicitly provides an estimate of the ET response to VPD,
given assumptions about ecosystem plant characteristics through the
parameters g$_1$ and uWUE. Equation \ref{d_et} is generic to external
environmental factors and the timescale over which a VPD response
acts, and has a straightforward conceptual meaning. It provides the ET
response, given other quantities and parameters held fixed. So, while
the plant parameters g$_1$ and uWUE may vary with plant species and
environmental conditions like soil moisture, we can still assess the
ET response to VPD for a given plant species, or for a given soil
moisture condition.

However, given that g$_1$ and uWUE can vary, and that they both have a
large impact on the ET response (Figure \ref{full}, \ref{sign}) it is
useful to examine how these quantities vary in previous literature,
framed by our results on ET responses. In \citet{Zhou_2015} most plant
types have uWUE similar to our median value, where whether ET
response is positive (water intensive) or negative (water conservative)
depends on VPD. However, crops, which we expect to prioritize carbon
uptake over water conservation, have a higher uWUE closer to our
maximum value. This higher uWUE results in a higher likelihood for an
increasing ET response to VPD, which intuitively makes sense given
that we expect crops to keep stomata open for access to carbon, at the
cost of water loss. Shrubs, and to a lesser extent grass, gave an uWUE
closer to the minimum uWUE. For these plant types, we would then
expect a higher likelihood of a water conservative response; that is, a
decrease of ET in response to VPD. It is important to note that, while
\citet{Zhou_2015} did not examine the roll of soil moisture in within
plant-type variability of uWUE, we might expect some variation,
especially in extreme cases when soil water becomes a limiting factor.

In \citet{Medlyn_2017} there is not as robust of a relationship
between plant type and g$_1$ as there was for uWUE in
\citet{Zhou_2015}. Given that there is a robust relationship between
g$_1$ and plant type in \citet{Lin_2015}, the ambiguous results in
\citet{Medlyn_2017} could be due to noise in ecosystem scale
observations, the consequences of imposing a model structure in some
of the g$_1$ observations, or real ecosystem-scale within-plant type
variability. These differences between leaf-scale behavior and
ecosystem scale behavior highlight the importance of understanding how
and why slope parameters vary at the ecosystem scale, as argued by
\citet{Medlyn_2017}. However, if we assume that there is some real
analogy between leaf scale results in \citet{Lin_2015} and ecosystem
scale behavior, and that some of the ambiguity in \citet{Medlyn_2017}
is due to model and/or observational error, then we can make some
connections between g$_1$-induced ET response variability and plant
characteristics, which are subject to the strong caveat that we still
do not fully understand the g$_1$'s utility at the ecosystem
scale. Extrapolating \citet{Lin_2015}'s results (e.g. Fig. 2) to the
ecosystem scale would reveal the following relationships between g$_1$
and plant types:

@@

Additionally, the sensitivity of ET response to plant parameter
highlights the importance of understanding and developing stomatal
conductance models and parameters for the ecosystem scale. In
particular, why and how g$_1$ behavior at the leaf and ecosystem scale
are not directly analogous much be understood. While the derivation
here was explicit about the assumptions of a constant g$_1$ term with
respect to the VPD perturbation, many sophisticated land surface
models and earth system models employing stomatal conductance models
make similar, and even stronger assumptions about the stationarity of
VPD slope terms. In particular, many models will assume a constant
g$_1$ term within a given plant functional type, and research derived
from these models will not acknowledge limitations of this assumption,
or difficulty in the literature with quantifying an ecosystem scale
slope parameter \citep{Medlyn_2017}. Given that something so
fundamental as the ET response to VPD varies strongly with g$_1$, we
must invest more in understanding g$_1$'s behavior at the ecosystem
scale, and the efficacy of models using g$_1$'s for ecosystem-scale
fluxes.

So far we have discussed the role of different parameters in the ET
response. However, the form of the ET response is imposed by our
choice of stomatal conductance model. We now examine how the model
choice, and specifically the exponent of the VPD dependence in the
stomatal conductance model, can impact the general form of the ET-VPD
relationship. This is relevant as many models include an
exponentiated VPD term \citep[e.g.,][]{Leuning_1990}, but the value
of the exponent can vary. It is useful to understand how generally the
choice of model imposes the functional form of the ET-VPD
relationship, given its fundamental importance for land atmosphere
interaction.
\subsection{Functional form of ET dependence on VPD and its relation
  to the VPD exponent}
\label{functional_form}

The theory described in Sect. \ref{sign_term} indicates that for a
given $uWUE$ and $g_1$, the ET dependence on VPD should be concave
upward, which is confirmed by eddy covariance data across most PFTs. In other words, there should be some local minimum in ET at a
critical VPD$_{crit}$, assuming the scaling and plant terms  (e.g. aerodynamic
conductance, $\Delta$, $g_1$ and $uWUE$) are held fixed. This result warrants further
investigation, because to our knowledge no one has derived the
theoretical ecosystem-scale relationship between ET and VPD while
controlling for other environmental conditions. In particular, from
personal communication, there is an apparent lack of consensus over
whether the shape of the ET-VPD curve should be concave upward (our
result) or concave downward in the absence of dramatic water stress. Given that understanding the ET-VPD
relationship of the one-way plant response is fundamental to
hypothesizing about any feedbacks between the land surface and the
atmosphere, we analyze why our derived ET-VPD relationship is concave
upward, particularly with respect to the exponent of VPD dependence in
$uWUE$ and the Medlyn unified stomatal conductance model.

>>>>

The Medlyn model adopted to ecosystem scale analysis in
\citet{Medlyn_2017}. In units of m s$^{-1}$, the ecosystem scale
stomatal conductance is:

@@

  where GPP is the ecosystem scale gross primary production, and $g_1$
  is an ecosystem scale analogue to $g_{1-leaf}$.

While Eq. (\ref{medlyn}) can be used in PM (Eq. (1)), it will make
analytical work with the function intractable because $GPP$ is
functionally related to ET itself. Additionally, a perturbation to VPD
should induce a physiological plant response that will alter GPP and
cause an indirect change in stomatal conductance, in addition to the
direct effect of VPD in Eq. (\ref{medlyn}). Therefore, in order to
derive the response of ET to VPD, we must account for the functional
relationship between GPP, ET, and VPD, and its effect on stomatal
conductance. We can use aforementioned semi-empirical results of
\citet{Zhou_2015} as a tool to approach this
problem. \citet{Zhou_2015}, showed that  underlying Water Use
Efficiency (uWUE):

@@

  is relatively constant across time and moisture conditions within a
  plant functional type, and correctly captures a constant
  relationship between GPP, ET and VPD over a diurnal cycle during the
  growing season \citep{Zhou_2014}. The theoretical derivation of the
  square root VPD dependence in $uWUE$ leverages the same assumptions
  used in \cite{MEDLYN_2011} to derive the square-root VPD dependence
  of the stomatal conductance model (Eq.  (\ref{medlyn})).  We can use
  uWUE to remove the $GPP$ dependence from $g_s$ in a way that makes
  PM analytically tractable:

@@

With Equation \ref{et} we can take the partial derivative with respect
to VPD:

@@

  This derivation relied either implicitly or explicitly on several
  assumptions. First, we assume that VPD at the leaf surface is the
  same as VPD at measurement height; physically this implies that
  leaves are perfectly coupled to the atmosphere. In reality, for some
  conditions and plant types the leaves can become decoupled from the
  boundary layer \citep{De_2017, Medlyn_2017}. Therefore, our
  derivation will be most applicable in times like the growing season
  (when we also expect uWUE to be most valid), when relatively high
  insolation induces instability and convective boundary layers, and
  we would expect the surface to be generally well coupled. An
  additional assumption in the formulation of $uWUE$ \citep{Zhou_2014,
    Zhou_2015} and \citet{Medlyn_2017}'s stomatal conductance model is
  that direct soil evaporation (E) contributions to ET remain small
  relative to transpiration (T). This should be more true during the
  growing season. The ratio of E to T may increase immediately after
  rainfall events due to high soil moisture and ponding, but VPD is
  generally low anyways during these times. However, some plant types
  allow for systematically larger contributions of E in ET,
  particularly those with sparse canopies and smaller relative amounts
  of transpiration. We therefore might expect that the theory will be
  most applicable to forest PFTs, which will be most strongly coupled
  to the boundary layer due to larger surface roughness, and will also
  generally have the highest ratios of transpiration to
  evaporation. Finally, we assume that $g_1$ and $uWUE$ are constant
  with respect to the conceptual VPD perturbation. Both quantities
  have been shown to be relatively constant with respect to changes in
  % TODO: make this a general discussion on partial derivatives and
  % what things are constant/how to interpret?
  VPD \citep{Franks_2017, Zhou_2014}. These parameters will however
  vary with plant species and characteristics \citep[e.g. wood density,
  ][]{Lin_2015}, as well as environmental conditions including soil
  water content and temperature.
  % TODO: write exact parameter ranges we use, and introduce
  % scaling/sign term here?
  % \item
  To account for both this variability in plant parameters and the
  environment, we will systematically analyze how the ET response to
  VPD (Equation \ref{d_et}) varies. Equation \ref{d_et} includes a
  ``sign'' term that determines the sign of the response in addition
  to magnitude:

@@

and a ``scaling'' term multiplying the ``sign'' term:

@@

In the ``sign term'' most of the quantities are relatively constant,
except for VPD and the plant parameters g$_1$ and uWUE. In the scaling
term, most of the terms are relatively constant with the exception of
aerodynamic conductance ($g_a$) and temperature (especially its effect
through $\Delta$). To determine the range of probable ET responses to
VPD we will systematically vary these parameters according to Table
\ref{param_varying}, while all other parameters are held fixed (Table
\ref{param_fixed}). Physical variables ($g_a$, $T$) and plant
physiological parameters (g$_1$, uWUE) are varied according to
literature-based expectations for a range of growing season conditions
and plant types \citep{Zhou_2015, Medlyn_2017}. Using this previous
literature we can connect the effect of varying plant parameters to
specific plant types and characteristics.

@@

\begin{table}
  \caption{Variable quantities in the ET response to VPD. Each value
    is varied to determine the effect of a range of expected plant and
    environmental conditions on ET response to VPD. A citation is
    provided in cases where the quantities are directly derived from
    previous literature. Conceptually, min. values are extracted from
    literature to correspond to approximately the 15th percentile of
    observed conditions during the growing season, med. values
    correspond to approximately the 50th percentile, and max. values
    correspond to approximately the 85th percentile.}
  \label{param_varying}
  \centering
  \begin{tabular}{l c c c c c}
    \hline
    Symbol & units [units in citation] & min  & med & max
    & citation  \\
    \hline
    \input{param_varying.tex}
    \hline
  \end{tabular}
\end{table}

\begin{table}
  \caption{Quantities that are fixed in the ET response to VPD
    (relative to those in Table \ref{param_varying}).}
  \label{param_fixed}
  \centering
  \begin{tabular}{l c c}
    \hline
    Symbol & units & value \\
    \hline
    \input{param_fixed.tex}
    \hline
  \end{tabular}
\end{table}

@@

\begin{figure}
  \centering \includegraphics{./fully_idealized.pdf}
  \caption{The functional form of $\frac{\partial \; ET}{\partial
      \; VPD}$ for minimum, median and maximum values of both the sign
    term and the scaling term.}
  \label{full}
\end{figure}

Examining the sign term independent of the scaling term illuminates
the role of the two plant parameters, g$_1$ and uWUE, in determining
the degree to which the response is water conservative
($\frac{\partial \; ET}{\partial \; VPD} < 0$) or water intensive
($\frac{\partial \; ET}{\partial \; VPD} > 0$) (Figure \ref{sign})
. Higher g$_1$ and uWUE shift the curve towards increasing ET
responses with VPD (water intensive), and smaller g$_1$ values lead to
a higher non-linear VPD dependence of the response.

\begin{figure}
  \centering \includegraphics{./fully_idealized_sign.pdf}
  \caption{The functional form of $\frac{\partial \; ET}{\partial
      \; VPD}$ evaluated at the median value of the scaling term, for
    varying values of g$_1$ and uWUE as given in Table \ref{param_varying}.}
  \label{sign}
\end{figure}

\begin{figure}
  \centering \includegraphics{./fully_idealized_scaling.pdf}
  \caption{The functional form of $\frac{\partial \; ET}{\partial
      \; VPD}$ evaluated at the median value of the sign term, for
    varying values of g$_a$ and T as given in Table \ref{param_varying}.}
  \label{scaling}
\end{figure}

Examining the scaling term independent of the sign term shows how
aerodynamic conductance (communication between the atmosphere and the
surface) and temperature (controls the efficiency of
energy conversion to latent heat through $\Delta$), amplify or suppress the plant
physiological response represented in the sign term (Figure \ref{scaing}). Both lower
temperatures and higher aerodynamic conductance lead to amplified ET
response to VPD, with the variability of aerodynamic conductance
resulting in a slightly higher variability of ET response relative to
temperature.

@@

Interpreting both the derivation and results of this manuscript hinges
on an understanding and appreciation for the usefulness of partial
derivatives for understanding the behavior of complex systems. Our
results are a conceptual tool for answering the question, ``What is
the ET response to VPD with all other quantities held fixed?'' This is
useful, and we would argue is critical, for understanding complex
responses when many quantities are varying simultaneously. If we
cannot understand the contribution of one term to observed
variability, independent of other confounding factors, there is little
hope of disentangling and understanding the relative role of many
quantities in determining observed variability.  Equation \ref{d_et}
explicitly provides an estimate of the ET response to VPD, given
assumptions about ecosystem plant characteristics through the
parameters g$_1$ and uWUE. Eq. (\ref{d_et}) is generic to external
environmental factors and the timescale over which a VPD response
acts, and has a straightforward conceptual meaning. It provides the ET
response, given other quantities and parameters held fixed, subject to
the assumptions outlined in Sect. \ref{methods}. So, while the plant
parameters g$_1$ and uWUE may vary with plant species and
environmental conditions like soil moisture, we can still assess the
ET response to VPD for a given plant species, or for a given soil
moisture condition.

However, given that g$_1$ and uWUE can vary, and they both have a
large impact on the ET response (Figure \ref{full}, \ref{sign}), it is
useful to examine how these quantities vary in previous literature,
framed by our results on $\frac{\partial \; ET}{\partial \; VPD}$. In
\citet{Zhou_2015} most plant types have uWUE similar to our median
value, under which the sign of the ET response depends on
VPD. However, crops, which we expect to prioritize carbon uptake over
water conservation, have a higher uWUE closer to our maximum
value. This higher uWUE results in a higher likelihood for an
increasing ET response to VPD, which intuitively makes sense given
that we expect crops to keep stomata open for access to carbon, at the
cost of increased water loss during high VPD. Shrubs, and to a lesser
extent grass, have an uWUE closer to the minimum uWUE. For these plant
types, we would then expect a higher likelihood of a water
conservative response; that is, a decrease of ET in response to
VPD. It is important to note that, while \citet{Zhou_2015} did not
examine the role of soil moisture for within plant-type variability of
uWUE, we might expect some variation, especially in extreme cases when
soil water becomes a limiting factor.

In \citet{Medlyn_2017} there is not as robust of a relationship
between plant type and g$_1$ as there was for uWUE in
\citet{Zhou_2015}. Given that there is a robust relationship between
g$_1$ and plant type in \citet{Lin_2015}, the ambiguous results in
\citet{Medlyn_2017} could be due to noise in ecosystem scale
observations, the consequences of imposing a model structure in some
of their g$_1$ esimations, or real ecosystem-scale within-plant type
variability. These differences between leaf-scale behavior and
ecosystem scale behavior highlight the importance of understanding how
and why slope parameters show increased variability at the ecosystem
scale, as argued by \citet{Medlyn_2017}. However, if we assume that
there is some real analogy between leaf scale results in
\citet{Lin_2015} and ecosystem scale behavior, and that some of the
ambiguity in \citet{Medlyn_2017} is due to model and/or observational
error, then we can make some connections between g$_1$-induced ET
response variability and plant characteristics, which are subject to
the strong caveat that we still do not fully understand g$_1$'s
utility at the ecosystem scale. Extrapolating \citet{Lin_2015}'s
results (e.g. Fig. 2) to the ecosystem scale would reveal the
following relationships between g$_1$ and plant types:

@@

Additionally, the sensitivity of ET response to plant parameters
highlights the importance of understanding and developing stomatal
conductance models and parameters for the ecosystem scale. In
particular, why and how g$_1$ behavior at the leaf and ecosystem scale
are not directly analogous much be understood. While the derivation
here was explicit about the assumption of a constant g$_1$ term with
respect to the VPD perturbation, many sophisticated land surface
models and earth system models employing stomatal conductance models
make similar, and even stronger assumptions about the stationarity of
VPD slope terms. Many models will assume a constant g$_1$ term within
a given plant functional type, and research derived from these models
often does not acknowledge limitations of this assumption, or the
difficulty in the literature with quantifying an ecosystem scale slope
parameter \citep{Medlyn_2017}. Given that something so fundamental as
the ET response to VPD varies strongly with g$_1$, we must invest more
in understanding g$_1$'s behavior at the ecosystem scale, and the
efficacy of models using g$_1$'s for ecosystem-scale
fluxes. Fortunately, our framework, and the novel general approach of
using robust ecosystem-scale ratios (uWUE) to remove the implicit
dependence of stomatal conductance on ET, is flexible enough to be
applied to any future-derived ecosystem-scale stomatal conductance
model that contains a dependence on GPP.

So far we have discussed the role of different parameters in
modulating the ET response. However, the form of the ET response is
imposed by our choice of stomatal conductance model. We now examine
how the model choice, and specifically the exponent of the VPD
dependence in the stomatal conductance model, can impact the general
form of the ET-VPD relationship. This is relevant as many models
include an exponentiated VPD term \citep[e.g.,][]{Leuning_1990}, but
the value of the exponent can vary. It is useful to understand how the
choice of model imposes the functional form of the ET-VPD
relationship, given its fundamental importance for land atmosphere
interaction.

\subsection{Functional form of ET dependence on VPD and its relation
  to the VPD exponent}
\label{functional_form}

The results presented in Sect. \ref{results} indicates that for a
given $uWUE$ and $g_1$, the ET dependence on VPD should be concave
upward. In other words, there should be some local minimum in ET at a
critical VPD$_{crit}$, assuming the scaling and plant terms  (e.g. aerodynamic
conductance, $\Delta$, $g_1$ and $uWUE$) are held fixed. This result warrants further
investigation, because to our knowledge no one has derived the
theoretical ecosystem-scale relationship between ET and VPD while
controlling for other environmental conditions. In particular, from
personal communication, there is an apparent lack of consensus over
whether the shape of the ET-VPD curve should be concave upward (our
result) or concave downward in the absence of dramatic water stress. Given that understanding the ET-VPD
relationship of the one-way plant response is fundamental to
hypothesizing about any feedbacks between the land surface and the
atmosphere, we analyze why our derived ET-VPD relationship is concave
upward, particularly with respect to the exponent of VPD dependence in
$uWUE$ and the Medlyn unified stomatal conductance model.

### 2019-02-27 doc/paper/conclusions.tex
about to start responding to pierre's comments

ET response to VPD can vary from strongly water conservative (ET
always decreasing in response to increasing VPD), to strongly water
intensive (ET always increasing in response to VPD), which is
indicative of the large observed diversity of stomatal regulation and
plant physiological strategy. Higher uWUE and g$_1$ values make the ET
response more likely to be positive, while decreasing temperature and
increasing g$_a$ amplify the magnitude of the response. Previous
literature \citep{Zhou_2015, Lin_2015} suggests that crops, through
association with higher g$_1$ and uWUE, are more likely to exhibit a
positive ET response to VPD. Shrubs (lower uWUE), C4 plants (lower
g$_1$), gymnosperm trees (lower g$_1$), and plants in a arctic and
boreal climates (lower g$_1$ are more likely to exhibit a negative ET
response to VPD. However interpretation of g$_1$-induced variability
is partially muddied by ambiguity in upscaling leaf-scale estimations
of g$_1$ to the ecosystem scale \citep{Medlyn_2017}.

@@

Our results are also applicable to understanding the impact of
expected increases in VPD induced by global change.  Plant
physiological responses to direct CO$_2$ effects
\citep[e.g.,][]{Swann_2016, Lemordant_2018} receives more attention
than physiological response to indirect effects like increased
VPD. Here, we provide a framework for understanding ET response to ET
using a simplified model of two plant parameters. Feedbacks between the
land and the atmosphere may alter the net response to a long-timescale
global VPD perturbation, but our focus on the one way plant response
to a VPD perturbation in the atmospheric boundary layer is an
important first step to disentangling such feedbacks, both in
observations and model simulations of the present and future. By
removing Penman Monteith's dependence on implicit relationships
between GPP, VPD, and ET, we allow for explicit, accurate future
analysis of plant-VPD feedbacks in the atmospheric boundary layer
(Eq. (\ref{full})). Our approach can be extended to examine varying
plant response to more nuanced consideration of plant type and
climate. Any plant physiological heterogeneity or feedback that can be
conceptualized with shifts in $g_1$ \citep[e.g.][]{Lin_2015,
Medlyn_2017} and/or uWUE \citep[e.g.][]{Zhou_2014} are representable
within our framework, which opens the door for a hierarchy of more
sophisticated climate- and plant-specific analysis of ET sensitivity
to environmental variables (including VPD). We argue that such
simplified conceptual frameworks are critical tools for disentangling
land-atmosphere feedbacks at various scales, from diurnal to seasonal
and beyond, and to characterize ET response in a warmer,
atmospherically drier, and enriched CO$_2$ world.

>>>>

ET response to VPD can vary from strongly water conservative (ET
always decreasing in response to increasing VPD), to strongly water
intensive (ET always increasing in response to VPD), which is
indicative of the diversity of possible plant water conservation
strategies. Higher uWUE and g$_1$ values increase the likelihood of a
postive ET response, while decreasing temperature and increasing
g$_a$ amplify the magnitude of the response. Previous literature
\citep{Zhou_2015, Lin_2015} suggests that crops, through association
with higher g$_1$ and uWUE, are more likely to exhibit a positive ET
response to VPD. Shrubs (lower uWUE), C4 plants (lower g$_1$),
gymnosperm trees (lower g$_1$), and plants in arctic and boreal
climates (lower g$_1$) are more likely to exhibit a negative ET
response to VPD. However interpretation of g$_1$-induced variability
is partially muddied by ambiguity in upscaling leaf-scale estimations
of g$_1$ to the ecosystem scale \citep{Medlyn_2017}.

@@

Our results are also applicable to understanding the impact of
expected increases in VPD induced by global change.  Plant
physiological responses to direct CO$_2$ effects
\citep[e.g.,][]{Swann_2016, Lemordant_2018} receives more attention
than physiological response to indirect effects like increased
VPD. Here, we provide a framework for understanding ET response to ET
using a simplified model of two plant parameters. Feedbacks between the
land and the atmosphere may alter the net response to a long-timescale
global VPD perturbation, but our focus on the one way plant response
to a VPD perturbation in the atmospheric boundary layer is an
important first step to disentangling such feedbacks, both in
observations and model simulations of the present and future. By
removing Penman Monteith's dependence on implicit relationships
between GPP, VPD, and ET, we allow for explicit, accurate future
analysis of plant-VPD feedbacks in the atmospheric boundary layer
(Eq. (\ref{et})). Our approach can be extended to examine varying
plant response to more nuanced consideration of plant type and
climate. Any plant physiological heterogeneity or feedback that can be
conceptualized with shifts in $g_1$ \citep[e.g.][]{Lin_2015,
Medlyn_2017} and/or uWUE \citep[e.g.][]{Zhou_2014} are representable
within our framework, which opens the door for a hierarchy of more
sophisticated climate- and plant-specific analysis of ET sensitivity
to environmental variables (including VPD). We argue that such
simplified conceptual frameworks are critical tools for disentangling
land-atmosphere feedbacks at various scales, from diurnal to seasonal
and beyond, and to characterize ET response in a warmer,
atmospherically drier, and enriched CO$_2$ world.
